Structuring Theories with Implicit Morphisms

### 2018-10-13 mmt.tex
typos

The only aspect of the inference system for these judgments that is relevant to our purposes here is the rules for constants:
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
Here, in order to avoid case distinctions for the case where a definition is present or not, we use a combined typing+equality judgment: $e_1=e_2:E$ represents the conjunction of $e_1:E$, $e_2:E$, and $e_1=e_2$.
Correspondingly, the definition of the homomorphic extension of a morphism with domain $S$ includes the following case for constants:
\[M(c) = \cas{e\mifc (c:E) \in \flt{S}, (c:=e)\in\flt{M} \\ M(e) \mifc (c:E=e)\in \flt{S}}\]
Here if $c$ has a definiens in $S$, we expand it before applying $M$.%
\footnote{The \mmt tool also $c:e\in \flt{M}$ even if $(c:E=e')\in\flt{S}$ has a definition. In that case, \mmt checks $\vdash_T M(e')=e$ and puts $M(c)=e$. This is important for efficiency but not essential for our purposes here.}

@@

In a theory declaration, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions only previously introduced constants.
We omit the remaining productions for expressions here, which allow forming complex expressions using application, binding, variables, literals, etc.
\end{definition}

@@

At this point, this grammar only allows forming flat theories, i.e., $\Thy=\thy$.
It is intentionally independent of the concrete choice of modularity principles, i.e., the kinds of module-structuring declarations (e.g., includes) and the module-expression-forming operators (e.g., union).
But it already introduces all non-terminals for modularity: new declaration kinds for structured theories arise by adding production for the non-terminal $Dec$ and new theory expressions are formed by adding productions for the non-terminal $T$.
This makes it easy to incrementally add new declaration kinds and operators to the language.

@@

\begin{definition}[Diagram]
A \textbf{diagram} is a list of theory and morphism declarations:
\begin{grammar}
Dia    & \rep{(TDec \alt MDec)}          & \text{diagrams}\\
\end{grammar}
A diagram is well-formed if each atomic theory/morphism declaration has a unique name is well-formed relative to the diagram preceding it.
\end{definition}

@@

\paragraph{Flattening}
Just like the syntax, the definition of flattening is \textbf{compositional} in the sense that new modularity principles can be added later independently of each other.
In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $flt{Y}$.

@@

The flat variant $\cn{Group}^\flat$ of our refactored theory $\cn{Group}$ from \autoref{syn:incl} is hence given by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup} except for global identifiers.
\end{example}

@@

\begin{example}[Union (continued from Ex.~\ref{sem:union})]\label{rel:union}
We have the following soundness theorem for unions: $S_i\harr S_1\cup S_2$.
Moreover, $S_1\cup S_2$ is a colimit in the category of theories and morphism, and $M_1\cup M_2\in\Mo{S_1\cup S_2}{T}$ is the universal morphism out of it.
\end{example}

>>>>

The only aspect of the inference system for these judgments that is relevant to our purposes here is the rules for constants:
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
Here, in order to avoid case distinctions for the case where a definition is present or not, we use a combined typing+equality judgment: $e_1=e_2:E$ represents the conjunction of $e_1:E$, $e_2:E$, and $e_1=e_2$.
Correspondingly, the definition of the homomorphic extension of a morphism with domain $S$ includes the following case for constants:
\[M(c) = \cas{e\mifc (c:E) \in \flt{S}, (c:=e)\in\flt{M} \\ M(e) \mifc (c:E=e)\in \flt{S}}\]
Here if $c$ has a definiens in $S$, we expand it before applying $M$.%
\footnote{The \mmt tool accepts $c:e\in \flt{M}$ even if $(c:E=e')\in\flt{S}$ has a definition. In that case, \mmt checks $\vdash_T M(e')=e$ and puts $M(c)=e$. This is important for efficiency but not essential for our purposes here.}

@@

In a theory declaration, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions over the previously introduced constants only.
We omit the remaining productions for expressions here, which allow forming complex expressions using application, binding, variables, literals, etc.
\end{definition}

@@

At this point, this grammar only allows forming flat theories, i.e., $\Thy=\thy$.
It is intentionally independent of the concrete choice of modularity principles, i.e., the kinds of module-structuring declarations (e.g., includes) and the module-expression-forming operators (e.g., union).
But it already introduces all non-terminals for modularity: new declaration kinds for structured theories arise by adding productions for the non-terminal $Dec$ and new theory expressions are formed by adding productions for the non-terminal $T$.
This makes it easy to incrementally add new declaration kinds and operators to the language.

@@

\begin{definition}[Diagram]
A \textbf{diagram} is a list of theory and morphism declarations:
\begin{grammar}
Dia    & \rep{(TDec \alt MDec)}          & \text{diagrams}\\
\end{grammar}
A diagram is well-formed if each atomic theory/morphism declaration has a unique name and is well-formed relative to the diagram preceding it.
\end{definition}

@@

\paragraph{Flattening}
Just like the syntax, the definition of flattening is \textbf{compositional} in the sense that new modularity principles can be added later independently of each other.
In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $\flt{Y}$.

@@

The flat variant $\cn{Group}^\flat$ of our refactored theory $\cn{Group}$ from \autoref{syn:incl} is hence given by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup} except for global identifiers from $\cn{Monoid}$ -- i.e. the prefixes $t$ in $t?n$.
\end{example}

@@

\begin{example}[Union (continued from Ex.~\ref{sem:union})]\label{rel:union}
We have the following soundness theorem for unions: $S_i\harr S_1\cup S_2$.
Moreover, $S_1\cup S_2$ is a colimit in the category of theories and morphism, and given $M_1\in\Mo(S_1,T)$ and $M2\in\Mo{S_2,T}$ we have the universal morphism $M_1\cup M_2\in\Mo{S_1\cup S_2}{T}$ out of it.
\end{example}
